Box 1 - TDT Curves

### 2023-09-27 ms/ms.qmd
Integrating Vanessa's comments. May need to edit last paragraph to reduce some duplication.

## Box 1 – Thermal Death Time framework case studies

Parameters from Thermal Death Time (TDT) curves provide a powerful tool for understanding the thermal sensitivity of organisms. Comparative approaches using such data can provide a way for us to evaluate differences in population and species sensitivity, and identify the key ecological and life-history traits that increase vulnerability to heat stress. Nonetheless, while static measures of thermal tolerance (e.g., $CT_{max}$) are widespread, parameters from TDT curves  are much less frequently estimated. In addition, we have a poor understanding of how dynamic (TDT) and static ($CT_{max}$) lethal measures correlate with each other and with sub-lethal measures of damage, such as critical fertility limits (CFLs). Identifying the correlation between lethal and sub-lethal limits can provide powerful ways to further clarify species vulnerability to heat stress that might lead to population extinction well before mortality is identified.

A number of significant challenges imped are ability to take the TDT and sub-lethal frameworks into a comparative context. First, ethical, conservation, and logistical challenges limit our ability to measure and estimate key parameters for many species, creating significant gaps in our knowledge across the tree of life. Second, while there has been a focus on species-level responses, population-level information is seldom incoperated into comparative frameworks limiting our ability to decompose sources of variance [@Pottier2023].

Powerful multivariate hierachical models (i.e., multi-response models) are capable of dealing with missing data, and provide a means by which multiple physiological thermal tolerance measures can be modelled simultaneously [@Noble2021; @Pottier2023]. When used in conjunction with an understanding of the phylogenetic relationships among species and/or populations, such modelling approaches can provide informed estimates for species missing data and explicitly estimate the covariance between tolerance measures while accounting for their uncertainty. Such insight can be used to inform policy and conservation-related decision making for species with little to no data.

@@

>>>>

## Box 1 – Comparing the utility of TDT Curves to other measures of heat tolerance within a comparative framework

Parameters from Thermal Death Time (TDT) curves provide a powerful tool for understanding the thermal sensitivity of organisms. But there are drawbacks to estimating TDT curves in comparison to other lethal (e.g. $CT_{max}$) and sublethal (CFL) measures. TDT curves are logistically challenging requiring more individuals for reliable estimates than other measures like $CT_{max}$ and CFL’s. TDT curves (and $CT_{max}$) also cannot be used for all organisms due to ethical and conservation considerations. Because of these limitations data on thermal sensitivity/heat vulnerability is missing for many species, creating significant knowledge gaps across the tree of life.

Phylogenetic comparative approaches could be leveraged to help fill these knowledge gaps because closely related species are often more likely to share similar traits owing to their evolutionary history (phylogenetic niche conservatism) (ref). If heat tolerance is linked to the phylogeny, shared characteristics that make species more or less heat tolerant will not be randomly distributed across the phylogeny but biased by species groups (ref). We can can use evolutionary relatedness to predict species thermal sensitivities and such insights can be used to inform policy and conservation related decisions making for species with little to no data.

While the comparative approach can help fill knowledge gaps it does not entirely solve the issue that TDT curves cannot be estimated for many organisms (logistical and ethical). $CT_{max}$ is less logistically challenging but also has ethical limitations. Ethical issues could be alleviated by the use of sublethal measures of thermal tolerance like CFL’s (ref), but we do not know how well lethal (TDT curves/$CT_{max}$) and sublethal (CFL’s) measures capture thermal sensitivities/vulnerabilities. Identifying whether a correlation exists between lethal and sub-lethal limits will help us understand the mechanisms of vulnerability to heat stress that might lead to population extinction well before mortality is identified. And if these traits are correlated different measures of thermal tolerance can be used for different organisms.

@@

Here we use powerful multivariate hierachical models (i.e., multi-response models) that are capable of dealing with missing data, and provide a means by which multiple physiological thermal tolerance measures can be modelled simultaneously [@Pottier2023; @Noble2021]. When used in conjunction with an understanding of the phylogenetic relationships among species and/or populations, such modelling approaches can provide informed estimates for species missing data and explicitly estimate the covariance between tolerance measures while accounting for their uncertainty.
